---
title: "Testing Gaussian GP No Inverse"
output: html_document
---
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\nc}[2]{\newcommand{#1}{#2}}
\nc{\vx}{\vect{x}}
\nc{\vmu}{\vect{\mu}}
\nc{\vSigma}{\vect{\Sigma}}
\nc{\vtheta}{\vect{\theta}}
\nc{\vr}{\vect{r}}
\nc{\vQ}{\vect{Q}}
\nc{\vgamma}{\vect{\gamma}}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(emdbook)
library(mcmc)
library(parallel)
library(MASS)
```

# Initialization of EP
First, we converts $\vmu_0$ and $\vSigma_0$, the moment parameters of the prior distribution, into natural parameters using the function `moment2natural()` 
```{r}
moment2natural <- function(mu, Sigma){
  Q <- chol2inv(chol(Sigma)) #solve(Sigma) # chol2inv(chol(Sigma)) might be better, but seems less stable?
  r <- Q %*% mu
  return(list(r=r, Q=Q))
}
```
which essentially computes
$$
\begin{align}
\vQ_0 &\leftarrow \vSigma_0^{-1} \\
\vr_0 &\leftarrow \vQ_0 \vmu_0
\end{align}
$$
Next, it sets the **global** _natural_ parameters to the prior natural parameters
$$
\begin{align}
\vQ &\leftarrow \vQ_0 \\
\vr &\leftarrow \vr_0
\end{align}
$$
And it sets the **global** _moment_ parameters to the prior moment parameters.
$$
\begin{align}
\vmu &\leftarrow \vmu_0 \\
\vSigma &\leftarrow \vSigma_0
\end{align}
$$
Finally, we set $\vr_k$ and $\vQ_k$ in each site $k=1, \ldots, \texttt{nsites}$ to be zeros. That is
$$
\vr_k = \begin{pmatrix}
0 \\
\vdots \\
0 
\end{pmatrix}\in\mathbb{R}^{\texttt{nparams}\times 1} \qquad \text{and} \qquad \vQ_k = \begin{pmatrix} 0 & \cdots & 0 \\ \vdots & \ddots & \vdots \\ 0 & \cdots & 0 \end{pmatrix}\in\mathbb{R}^{\texttt{nparams}\times \texttt{nparams}}
$$

```{r}
initialize_ep <- function(nsites, mu_prior, Sigma_prior){
  # Find dimensionality of parameters
  nparams <- nrow(Sigma_prior)
  # Convert mu_0 and Sigma_0 to their natural parameter versions
  natural_prior <- moment2natural(mu_prior, Sigma_prior)
  
  ep <- list(
    r=natural_prior$r, # Set global r and global Q to r_prior and q_prior. 
    Q=natural_prior$Q,
    mu=mu_prior,       # Set global mu and Sigma to mu_prior and Sigma_prior
    Sigma=Sigma_prior,
    r_list=array(0, dim=c(nparams, 1, nsites)),  # Store arrays containing r_i and Q_i for every site
    Q_list=array(0, dim=c(nparams, nparams, nsites))
  )
  return(ep)
}
```

# Site Update
The site update at site $\texttt{site_index}$ is mainly performed by the function `site_update`, shown below.

```{r}
site_update <- function(ep, site_index, pass_index, nsamples, local_likelihood){
  # Compute mu, Sigma of cavity distribution
  # moment_cavity <- moments_cavity(ep, site_index)
  # Sigma_c <- moment_cavity$Sigma
  # mu_c    <- moment_cavity$mu
  Q_c <- ep$Q - ep$Q_list[,,site_index]
  r_c <- ep$r - matrix(ep$r_list[,,site_index])
  # Check for positive definiteness of Sigma, otherwise, skip this
  if (is_positive_definite(Sigma_c)){
    # By sampling tilted, find new global natural parameters
    # rq_new <- sample_moments_tilted(site_index, local_likelihood, mu_c, Sigma_c, nsamples)
    rq_new <- sample_moments_tilted(site_index, local_likelihood, r_c, Q_c, nsamples)
    # Find new local natural parameters by subtracting natural params of cavity distribution
    r_i_new <- rq_new$r - ep$r + matrix(ep$r_list[,,site_index])
    q_i_new <- rq_new$Q - ep$Q + ep$Q_list[,,site_index]
    # Update EP local parameters
    # Return the new local natural parameter so we can do updates in a parallel fashion
    return(list(r_i=r_i_new, 
                q_i=q_i_new))
  } else {
    cat("Skipping site", site_index, "in pass", pass_index, "as Sigma of Cavity is not positive definite.\n")
    #TODO: Do we keep the old ones, or pass empty ones?
    return(list(
      r_i = matrix(ep$r_list[,,site_index]),
      q_i = ep$Q_list[,,site_index]
    ))
  }
}
```

### Moment Parameters of Cavity Distribution
It starts by calculating the **moment** parameters of the cavity distribution using the function `moments_cavity`, which takes an `ep` object, a site index and it outputs a list containing the **moment** parameters

```{r}
moments_cavity <- function(ep, site_index){
  # Subtract Q_i from Q_global, and r_i from r_global to find natural parameters
  # of cavity distribution
  Q_c <- ep$Q - ep$Q_list[,,site_index]
  r_c <- ep$r - matrix(ep$r_list[,,site_index])
  moment_cavity <- natural2moment(r_c, Q_c)
  return(list(mu_c=moment_cavity$mu, 
              Sigma_c=moment_cavity$Sigma))
}
```

These are calculated by subtracting the natural parameters of the $\texttt{site_index}^{\text{th}}$ approximation term from the current global natural parameters
$$
\begin{align}
\vQ_{c} &\leftarrow \vQ - \vQ_k \\
\vr_{c} &\leftarrow \vr - \vr_k
\end{align}
$$
and then the natural parameters are converted to moment parameters using the function `natural2moment()`
```{r}
natural2moment <- function(r, Q){
  Sigma <- chol2inv(chol(Q)) #solve(Q) #chol2inv(chol(Q))
  mu <- Sigma %*% r
  return(list(mu=mu, Sigma=Sigma))
}
```
using the formulas
$$
\begin{align}
\vSigma_c &\leftarrow \vQ_c^{-1} \\
\vmu_c &\leftarrow \vSigma_c \vr_c
\end{align}
$$

### Moment-Matching
After checking that $\vSigma_c$ is positive definite by looking at its eigenvalues
```{r}
is_positive_definite <- function(Sigma, tol=1e-8){
  # Compute eigenvalues
  eigenvalues <- eigen(Sigma, only.values = TRUE)$values
  # If smaller then threshold, set to zero
  eigenvalues[abs(eigenvalues) < tol] <- 0.0
  # If any is <= 0 then it's not positive definite
  return(ifelse(any(eigenvalues <= 0), FALSE, TRUE))
}
```
we proceed by getting **sample estimates** of the moment parameters using the function `sample_moments_tilted`
```{r}
sample_moments_tilted <- function(site_index, local_likelihood, r_c, Q_c, nsamples){
  # Create wrapper for tilted distribution. Will be used to sample from it
  #log_target <- function(x) log(local_likelihood(x)) + dmvnorm(x, c(mu_c), Sigma_c, log=TRUE)
  log_target <- function(x) {
    x_vector <- matrix(x)
    like_value <- log(local_likelihood(x)) 
    cavity_value <- (-0.5*t(x_vector)%*%(Q_c %*% x_vector) + t(r_c) %*% x_vector)[1]
    return(like_value + cavity_value)
  }
  # Sample from the targe distribution. First, find the mode and start sampling from there
  #sol <- optim(rep(0, nrow(Sigma_c)), function(x) -log(target(x)), method="BFGS", hessian=TRUE)
  sol <- optim(c(mu_c), function(x) -log_target(x), method="BFGS", hessian=TRUE)
  inv_hessian <- chol2inv(chol(sol$hessian))
  #samples <- metrop(function(x) log(target(x)), sol$par, nbatch=nsamples)$batch
  samples <- rwmh_log(start=sol$par, niter=nsamples, logtarget=log_target, Sigma=inv_hessian)
  # Compute mean and covariance of the samples, assign them to the new global moment parameters
  mu_global_new    <- matrix(apply(samples, 2, mean))
  Sigma_global_new <- cov(samples)
  # Convert new global moment parameters to natural parameters
  natural_global_new <- moment2natural(mu_global_new, Sigma_global_new)
  return(list(Q=natural_global_new$Q, r=natural_global_new$r))
}
```
This function samples $\widetilde{\boldsymbol{\mathbf{\gamma}}}_k^{(1)}, \ldots, \widetilde{\boldsymbol{\mathbf{\gamma}}}_k^{(N)}$ from
$$
f_k(\vx; \vgamma_k) \,\mathcal{N}(\vx; \vmu_c, \vSigma_c)
$$
using the Random-Walk Metropolis-Hastings algorithm provided by the `mcmc` package. We start the algorithm at the mode of the tilted distribution, which is found with `optim()` on minus the negative logarithm of the tilted distribution.

Once we've obtained the samples we calculate their mean and their variance-coviariance matrix. We take these sample estimates as the new global moment parameters
$$
\begin{align}
\boldsymbol{\mathbf{\mu}}^{\text{new}, k}&\leftarrow \frac{1}{N}\sum_{j=1}^N \widetilde{\boldsymbol{\mathbf{\gamma}}}_i^{(j)} \\
\boldsymbol{\mathbf{\Sigma}}^{\text{new}, k} &\leftarrow \frac{1}{N - 1}\sum_{j=1}^N \left(\widetilde{\boldsymbol{\mathbf{\gamma}}}_i^{(j)} - \boldsymbol{\mathbf{\mu}}^{\text{new}, k}\right) \left(\widetilde{\boldsymbol{\mathbf{\gamma}}}_i^{(j)} - \boldsymbol{\mathbf{\mu}}^{\text{new}, k}\right)^\top
\end{align}
$$
Finally, we convert these to natural parameters for the global approximation using `moment2natural()` as before.
$$
\begin{align}
\boldsymbol{\mathbf{Q}}^{\text{new}, k} &\leftarrow \left(\boldsymbol{\mathbf{\Sigma}}^{\text{new}, k}\right)^{-1} \\
\boldsymbol{\mathbf{r}}^{\text{new}, k} &\leftarrow \boldsymbol{\mathbf{Q}}^{\text{new}, k} \boldsymbol{\mathbf{\mu}}^{\text{new}, k}
\end{align}
$$

### Finding the New Site Natural Parameters
The last step is to subtract the cavity parameters from the newly-found global natural parameters using the formulas
$$
\begin{align}
\vr_k^{\text{new}} &\longleftarrow \vr^{\text{new}, k} - \vr_c = \vr^{\text{new}, k} - (\vr - \vr_k) = \vr^{\text{new}, k} - \vr  +\vr_k \\
\vQ_k^{\text{new}} &\longleftarrow \vQ^{\text{new}, k} - \vQ_c= \vQ^{\text{new}, k} - (\vQ - \vQ_k) = \vQ^{\text{new}, k}  - \vQ + \vQ_k
\end{align}
$$

### In case Cavity Variance-Covariance Matrix is not Positive Definite
If the variance coviariance matrix of the cavity distribution, $\vSigma_c$, is not positive definite then we cannot sample from the tilted distribution, because the cavity is undefined. In that case we simply set the new natural local parameters to be the same as before
$$
\begin{align}
\vr^{\text{new}}_k &\leftarrow \vr_k \\
\vQ^{\text{new}}_k &\leftarrow \vQ_k
\end{align}
$$

# Main Expectation Propagation Algorithm
### Summary of the Algorithm
The algorithm proceeds as follows

1. Initializes the parameters as seen above.
2. Cycles through $\texttt{npasses}$ passes. Each pass is an update to _every_ site, where it computes the update as seen before.
3. After updating all the sites and having obtained $\vr_1^{\text{new}},\ldots, \vr_{\texttt{nsites}}^{\text{new}}$ and $\vQ_1^{\text{new}}, \ldots, \vQ_{\texttt{nsites}}^{\text{new}}$ we obtain the new global natural parameters by summing all the new local natural parameters
$$
\begin{align}
\vQ^{\text{new}} &\leftarrow \sum_{k=1}^{\texttt{npasses}} \vQ_k^{\text{new}} \\
\vr^{\text{new}} &\leftarrow \sum_{k=1}^{\texttt{npasses}} \vr_k^{\text{new}}
\end{align}
$$

and finally it updates the values of the local natural parameters stored in the `ep` object.

### Damping for Stability
Having obtained $\vr^{\text{new}}$ and $\vQ^{\text{new}}$ we can use a damping factor $\alpha$, which is set to $\alpha=1.0$ if we wish to not use it, to slow down convergence while at the same time improving stability. Damping is performed as
$$
\begin{align}
\vQ^{\text{new, damped}} &\leftarrow \alpha\vQ^{\text{new}} + (1 - \alpha)\vQ \\
\vr^{\text{new, damped}} &\leftarrow \alpha\vQ^{\text{new}} + (1 - \alpha)\vr
\end{align}
$$
At the end, we simply convert the global natural parameters to the moment ones.

```{r}
ep_algorithm <- function(nsites, nparams, nsamples, npasses, mu_prior, Sigma_prior, local_likelihood, alpha=1.0){
  # Store history
  history_mu <- array(0, dim=c(nparams, 1, npasses))
  history_Sigma <- array(0, dim=c(nparams, nparams, npasses))
  # Store prior natural parameters, arrays of Q_i and r_i, global Q and r
  ep <- initialize_ep(nsites, mu_prior, Sigma_prior)
  # Each pass computes updates in parallel
  for (pass_index in 1:npasses){
    cat("Pass", pass_index,": Global r is\n", ep$r, "\nGlobal Q is\n", ep$Q, "\n")
    # Create a wrapper function so that we can use mclapply
    site_update_wrapper <- function(site_index){
      return(site_update(ep, site_index, pass_index, nsamples, local_likelihood))
    }
    # Find (mu, Sigma) of cavity distribution, construct tilted, sample from it 
    # and convert moment parameters to natural parameters.
    # updates <- list()
    # for (site in 1:nsites){
    #   updates[[site]] <- site_update_wrapper(site)
    # }
    updates <- mclapply(1:nsites, site_update_wrapper, mc.cores = 4)
    # Sum up all the updates and update ep local natural parameters
    r_new <- 0.0
    Q_new <- 0.0
    for (site in 1:nsites){
      ep$r_list[,,site] <- updates[[site]]$r_i
      ep$Q_list[,,site] <- updates[[site]]$q_i
      r_new <- r_new + updates[[site]]$r_i
      Q_new <- Q_new + updates[[site]]$q_i
    }
    # Update the global parameters with some damping 
    ep$r <- alpha*r_new + (1 - alpha)*(ep$r)
    ep$Q <- alpha*Q_new + (1 - alpha)*(ep$Q)
    # Store history 
    history_mu[,,pass_index] <- ep$mu
    history_Sigma[,,pass_index] <- ep$Sigma
  }
  # Update the global mu and Sigma
  global_moments <- natural2moment(ep$r, ep$Q)
  ep$mu <- global_moments$mu
  ep$Sigma <- global_moments$Sigma
  return(list(ep=ep, history_mu=history_mu, history_Sigma=history_Sigma))
}
```


# Example
Use a normal likelihood
```{r}
local_likelihood <- function(x){
  return(dmvnorm(x, mu=c(1, 1), Sigma=diag(2)))
}
```
Specify settings here
```{r}
nsites <- 20
nparams <- 2
nsamples <- 1000
npasses <- 5
mu_prior <- matrix(c(0,0))
Sigma_prior <- diag(2)
alpha <- 1.0
```

```{r}
a <- ep_algorithm(nsites, nparams, nsamples, npasses, mu_prior, Sigma_prior, local_likelihood, alpha)
```

# Run Example
### Initialize
```{r}
ep <- initialize_ep(nsites, mu_prior, Sigma_prior)
```

### Set pass index
```{r}
pass_index <- 1
```

### Set site index
```{r}
site_index <- 1
site <- site_index
```



# MY metrop
```{r}
library(MASS)
rwmh_log <- function(start, niter, logtarget, Sigma=NULL){
    # Starting Points and Their Density
    z <- start
    pz <- logtarget(z)
    # grab the dimensionality of z
    if (is.null(nrow(z))){
      d <- length(z)
    } else {
      d <- nrow(z)
    }
    # Store Samples
    samples <- matrix(0, nrow=niter, ncol=d)
    samples[1, ] <- z
    # Uniform Random Numbers
    log_u <- log(runif(niter))
    # Proposal: Normal Distribution
    if (is.null(Sigma)){
      Sigma <- diag(d)
    }
    normal_shift <- mvrnorm(n=niter, mu=rep(0, d), Sigma=Sigma)
    for (i in 2:niter){
        # Sample Candidates
        candidate <- z + normal_shift[i, ]
        # Candidate Densities
        p_candidate <- logtarget(candidate)
        # Accept/Reject for Chain 1
        if (log_u[i] <= p_candidate - pz){
          z <- candidate
          pz <- p_candidate
        }
        # Store samples
        samples[i, ] <- z
    }
    return(samples)
}
```




# Minimal Working Example
### Why Product of Gaussians is not Sampled Properly?
```{r}
mu1    <- c(0,0)
Sigma1 <- diag(2)
mu2    <- c(1,1)
Sigma2 <- diag(2)
normal1 <- function(x) dmvnorm(x, mu=mu1, Sigma=Sigma1)
normal2 <- function(x) dmvnorm(x, mu=mu2, Sigma=Sigma2)
normal_product <- function(x) normal1(x) * normal2(x)
```
Sample from the one centered at $(0, 0)$ and find its mean and variance-covariance matrix
```{r}
n1_samples <- rwmh_log(c(0,0), 10000, function(x) log(normal1(x)))
apply(n1_samples, 2, mean)
```

```{r}
cov(n1_samples)
```

Now do the same but with the product
```{r}
np_samples <- rwmh_log(c(0,0), 10000, function(x) log(normal_product(x)))
apply(np_samples, 2, mean)
```
Look at its variance-covariance matrix
```{r}
cov(np_samples)
```


# Plotting Contours
```{r}
ncontour <- 100
vals <- seq(-2, 2, length.out=ncontour)
grid <- meshgrid(vals)
vals_matrix <- cbind(matrix(c(grid[[1]])), matrix(c(grid[[2]])))
contour(x=vals, y=vals, z=matrix(normal_product(vals_matrix), nrow=ncontour))
contour(x=vals, y=vals, z=matrix(normal1(vals_matrix), nrow=ncontour))
```



# Should I draw from each one separately, multiply and then average?
```{r}
n1_samples <- rwmh_log(c(0,0), 2000, function(x) log(normal1(x)))
n2_samples <- rwmh_log(c(0,0), 2000, function(x) log(normal2(x)))
prod <- n1_samples * n2_samples
```

Nope, doesn't work cause I get 
```{r}
apply(prod, 2, mean)
```


